---
title: "Lab 1"
author: "Sofia Garrick 04/12/20"
output: pdf_document
---


# Version

```{r, echo=TRUE}
version
```

# Load Libraries

I decided to load `RColorBrewer` and some important GIS libraries from Chapter 2.

```{r, echo= TRUE, eval= FALSE}
library(sf)
library(raster)
library(sp)
```

# Coding Examples

### Example 1

Here is code I used to answer question 4 from the Chapter 2.6 exercises 
```{r, echo= TRUE, eval= TRUE}
library(sp) #reloading libraries since eval= FALSE above
library(raster)
my_raster_2 = raster(nrows = 10, ncols = 10, res = 1, xmn = 0, xmx = 10, ymn = 0, ymx = 10, vals = runif(100,0,10))
plot(my_raster_2, axes = TRUE)

```

### Example 2

I was interested in getting the specific HEX color codes from the RColorBrewer collections, so here is the code I used to determine the colors from the PuRd collection which is adapted from the code provided on this [web page](https://www.stat.ubc.ca/~jenny/STAT545A/block14_colors.html).
```{r, echo= TRUE, eval= TRUE}
library(RColorBrewer)
colors <- brewer.pal(n = 8, name = "PuRd")
colors
```
