---
title: "Lab 6"
author: "Sofia Garrick"
date: "5/26/2020"
output: html_document
---

# Source Information
The inputs to OAC are sourced entirely from the UK 2011 census, and are organised around three domains; demographic, housing and socio-economics. 

# Data 

### Load Data
```{r}
load("/Users/sofiagarrick/Documents/COLLEGE/GIS/GIS 3/lab 6/github/urban_analytics/10_Data_Reduction_Geodemographics/data/census_2011_UK_OA.RData")
```


### Subset Data to include only Liverpool
```{r}
Census_2011_Count <- merge(Liverpool,Census_2011_Count_All,by="OA",all.x=TRUE)
```

# Normalizing

### Caluclating Numerators
```{r}
OAC_Input <- as.data.frame(Census_2011_Count$OA)
colnames(OAC_Input) <- "OA"
# Loop through each row in the OAC input table
for (n in 1:nrow(OAC_Input_Lookup)){
      # Get the variables to aggregate for the row specified by n
      select_vars <- OAC_Input_Lookup[n,"England_Wales"]
      
      # Create a list of the variables to select
      select_vars <- unlist(strsplit(paste(select_vars),","))
      
      # Create variable name
      vname <- OAC_Input_Lookup[n,"VariableCode"] 
      
      # Creates a sum of the census variables for each Output Area
      tmp <- data.frame(rowSums(Census_2011_Count[,select_vars, drop=FALSE]))
      colnames(tmp) <- vname
      
      # Append new variable to the OAC_Input object
      OAC_Input <- cbind(OAC_Input,tmp)
      
      # Remove temporary objects
      remove(list = c("vname","tmp"))
} # END: Loop through each row in the OAC input table
```


### Remove attributes for Standardized Illness Ratio as it is being calculated later
```{r}
OAC_Input$k035 <- NULL
```


### Calculating Denominators
```{r}
OAC_Input_den <- as.data.frame(Census_2011_Count$OA)
colnames(OAC_Input_den) <- "OA"
# Create a list of unique denominators
den_list <- unique(OAC_Input_Lookup[,"Denominator"])
den_list <- paste(den_list[den_list != ""])
# Select denominators
OAC_Input_den <- Census_2011_Count[,c("OA",den_list)]
```


### Merge Numerators w/ Denominators
```{r}
OAC_Input <- merge(OAC_Input,OAC_Input_den, by="OA")
```


## Calculate percentages

### Ensure correct denominator w/ no ratios
```{r}
# Get numerator denominator list where the Type is "Count" - i.e. not ratio
K_Var <- OAC_Input_Lookup[OAC_Input_Lookup$Type == "Count",c(1,3)]
# View top 6 rows
head(K_Var)
```


```{r}
# Create an OA list / data frame
OAC_Input_PCT_RATIO <- subset(OAC_Input, select = "OA")
# Loop
for (n in 1:nrow(K_Var)){
  
  num <- paste(K_Var[n,"VariableCode"]) # Get numerator name
  den <- paste(K_Var[n,"Denominator"]) # Get denominator name
  tmp <- data.frame(OAC_Input[,num] / OAC_Input[,den] * 100) # Calculate percentages
  colnames(tmp) <- num
  OAC_Input_PCT_RATIO <- cbind(OAC_Input_PCT_RATIO,tmp) # Append the percentages
  
  # Remove temporary objects
  remove(list = c("tmp","num","den"))
}
```


### Join final two variables from original census table
```{r}
#Extract Variable
tmp <- Census_2011_Count[,c("OA","KS101EW0008")]
colnames(tmp) <- c("OA","k007")
#Merge
OAC_Input_PCT_RATIO <- merge(OAC_Input_PCT_RATIO,tmp,by="OA")
```

### Now calculate Standardized Illness Rate (k035) in Liverpool
```{r}
# Calculate rates of ill people 15 or less and greater than or equal to 65
ill_16_64 <- rowSums(Census_2011_Count[,c("KS301EW0005","KS301EW0006")]) # Ill people 16-64
ill_total <-   rowSums(Census_2011_Count[,c("KS301EW0002","KS301EW0003")]) # All ill people
ill_L15_G65 <- ill_total - ill_16_64 # Ill people 15 or less and greater than or equal to 65
# Calculate total people 15 or less and greater than or equal to 65
t_pop_16_64 <- rowSums(Census_2011_Count[,c("KS102EW0007","KS102EW0008","KS102EW0009","KS102EW0010","KS102EW0011","KS102EW0012","KS102EW0013")]) # People 16-64
t_pop <- Census_2011_Count$KS101EW0001 # All people
t_pop_L15_G65 <- t_pop - t_pop_16_64 # All people 15 or less and greater than or equal to 65
# Calculate expected rate
ex_ill_16_64 <- t_pop_16_64 * (sum(ill_16_64)/sum(t_pop_16_64)) # Expected ill 16-64
ex_ill_L15_G65 <- t_pop_L15_G65 * (sum(ill_L15_G65)/sum(t_pop_L15_G65)) # Expected ill people 15 or less and greater than or equal to 65
ex_ill <- ex_ill_16_64 + ex_ill_L15_G65 # total expected ill people
# Ratio
SIR <- as.data.frame(ill_total / ex_ill * 100) # ratio between ill people and expected ill people
colnames(SIR) <- "k035"
# Merge data
OAC_Input_PCT_RATIO <- cbind(OAC_Input_PCT_RATIO,SIR)
# Remove unwanted objects
remove(list=c("SIR","ill_16_64","ill_total","ill_L15_G65","t_pop_16_64","t_pop","t_pop_L15_G65","ex_ill_16_64","ex_ill_L15_G65","ex_ill"))
```


### Apply standardization & normalization procedures to input Data
```{r}
# Calculate inverse hyperbolic sine
OAC_Input_PCT_RATIO_IHS <- log(OAC_Input_PCT_RATIO[,2:61]+sqrt(OAC_Input_PCT_RATIO[,2:61]^2+1))
# Calculate Range
range_01 <- function(x){(x-min(x))/(max(x)-min(x))} # range function
OAC_Input_PCT_RATIO_IHS_01 <- apply(OAC_Input_PCT_RATIO_IHS, 2, range_01) # apply range function to columns
# Add the OA codes back onto the data frame as row names
rownames(OAC_Input_PCT_RATIO_IHS_01) <- OAC_Input_PCT_RATIO$OA
```

# Estimating the number of clusters

### Calculate total within sum of squares to determine appropirate # of clusters that are representative of the data
```{r}
library(ggplot2)
# Create a new empty numeric object to store the wss results
wss <- numeric()
# Run k means for 2-12 clusters and store the wss results
for (i in 2:12) wss[i] <- sum(kmeans(OAC_Input_PCT_RATIO_IHS_01, centers=i,nstart=20)$withinss)
# Create a data frame with the results, adding a further column for the cluster number
wss <- data.frame(2:12,wss[-1])
# Plot the results
names(wss) <- c("k","Twss")
ggplot(data=wss, aes(x= k, y=Twss)) + geom_path() + geom_point() + scale_x_continuous(breaks=2:12) + labs(y = "Total within sum of squares")
```
No large decreases but minor moderation of decrease around 7 or 8 clusters --> 7 cluster solution is chosen. 

# Building the geodemographic

### Cluster Analysis: load pre-run object that optimizes the partitioning of the clusers
```{r}
# Load cluster object
load("/Users/sofiagarrick/Documents/COLLEGE/GIS/GIS 3/lab 6/github/urban_analytics/10_Data_Reduction_Geodemographics/data/cluster_7.Rdata")
#code that is commented out is the object we are loading --> which is an algorithm that analyzes the partitioning of clusters but takes a while to run as it has a very large iteration to ensure stochasticity
#cluster_7 <- kmeans(x=OAC_Input_PCT_RATIO_IHS_01, centers=7, iter.max=1000000, nstart=10000)
```

### List of different outputs related to the cluster analysis
```{r}
# Show object content
str(cluster_7)
```


### Cluster Results
```{r}
# Lookup Table
lookup <- data.frame(cluster_7$cluster)
# Add OA codes
lookup$OA <- rownames(lookup)
colnames(lookup) <- c("K_7","OA")
# Recode clusters as letter
lookup$SUPER <- LETTERS[lookup$K_7]
```

### View Distribution of Results
```{r}
table(lookup$K_7)
```

# Map Results of Cluster Analysis

```{r message=FALSE, warning=FALSE}
# Load packages
library(rgdal)
library(tmap)
# Import OA boundaries
liverpool_SP <- readOGR("/Users/sofiagarrick/Documents/COLLEGE/GIS/GIS 3/lab 6/github/urban_analytics/10_Data_Reduction_Geodemographics/data/Liverpool_OA_2011.geojson", "Liverpool_OA_2011")

# Merge lookup
liverpool_SP <- merge(liverpool_SP, lookup, by.x="oa_code",by.y="OA")
m <- tm_shape(liverpool_SP, projection=27700) +
    tm_polygons(col="SUPER", border.col = "grey50",   palette="Set1",border.alpha = .3, title="Cluster", showNA=FALSE) +
  tm_layout(legend.position = c("left", "bottom"), frame = FALSE)

#Create leaflet plot
tmap_leaflet(m)
```

# Creating cluster descriptions and profiles

### Create Table of Index Scores
```{r}
# Merge Original Data (inc. denominators)
LiVOAC_Lookup_Input <- merge(lookup,OAC_Input,by="OA",all.x=TRUE)
# Remove Ratio Variables
LiVOAC_Lookup_Input$k007 <- NULL
LiVOAC_Lookup_Input$k035 <- NULL
# Create Aggregations by SuperGroup
SuperGroup <-aggregate(LiVOAC_Lookup_Input[,4:78], by=list(LiVOAC_Lookup_Input$SUPER),  FUN=sum)
# Create a data frame that will be used to append the index scores
G_Index <- data.frame(SUPER=LETTERS[1:7])
# Loop
for (n in 1:nrow(K_Var)){
  
  num <- paste(K_Var[n,"VariableCode"]) # Get numerator name
  den <- paste(K_Var[n,"Denominator"]) # Get denominator name
  tmp <- data.frame(round((SuperGroup[,num] / SuperGroup[,den]) / (sum(SuperGroup[,num])/sum(SuperGroup[,den]))*100)) # Calculate index score - these are also rounded
  colnames(tmp) <- num
  
  G_Index <- cbind(G_Index,tmp) # Append the index calculations
  
  # Remove temporary objects
  remove(list = c("tmp","num","den"))
}
# View the index scores
G_Index
```

## Visualize the Index Scores in a colorcoded table

```{r}
# Convert from wide to narrow format
G_Index_Melt <- reshape2::melt(G_Index, id.vars="SUPER")
# View the top of the new narrow formatted data frame
head(G_Index_Melt)
```

### Add shortened descriptions of variables
```{r}
# Recode the index scores into aggregate groupings
G_Index_Melt$band <- ifelse(G_Index_Melt$value <= 80,"< 80",ifelse(G_Index_Melt$value > 80 & G_Index_Melt$value <= 120,"80-120",">120"))
# Add a column with short descriptions of the variables
short <- read.csv("/Users/sofiagarrick/Documents/COLLEGE/GIS/GIS 3/lab 6/github/urban_analytics/10_Data_Reduction_Geodemographics/data/OAC_Input_Lookup_short_labels.csv")
G_Index_Melt <- merge(G_Index_Melt,short,by.x="variable",by.y="VariableCode",all.x=TRUE)
# Order the created factors appropriately - needed to ensure the legend and axis make sense in ggolot2
G_Index_Melt$band <- factor(G_Index_Melt$band, levels = c("< 80","80-120",">120"))
G_Index_Melt$VariableDescription <- factor(G_Index_Melt$VariableDescription, levels = short$VariableDescription)
```


### Visualize
```{r fig.width=7, fig.height=10}
library(ggplot2)
p <- ggplot(G_Index_Melt, aes(x=SUPER, y=VariableDescription, label=value, fill=band)) + 
  scale_fill_manual(name = "Band",values = c("#EB753B","#F7D865","#B3D09F")) +
  scale_x_discrete(position = "top") +
  geom_tile(alpha=0.8) +
  geom_text(colour="black")
p
```

### update cluster descriptions based on table

```{r}
lookup$Description <- LETTERS[lookup$K_7]

lookup$Description <- gsub("A", "Older, retired, flats", lookup$Description)
lookup$Description <- gsub("B", "Middle aged to older, no children or non-dependent children, house & car owners", lookup$Description)
lookup$Description <- gsub("C", "Younger, city dwelling families", lookup$Description)
lookup$Description <- gsub("D", "Younger city dwelling individuals working in Manufacturing, Utilities, and Warehouse", lookup$Description)
lookup$Description <- gsub("E", "Younger individuals, multicultural, city communities", lookup$Description)
lookup$Description <- gsub("F", "Older individuals, multicultural, city communities", lookup$Description)
lookup$Description <- gsub("G", "Young families, multicultural, city communities", lookup$Description)
```

# Create New Map w/ updated cluster descriptions

```{r message=FALSE, warning=FALSE}
# Load packages
library(rgdal)
library(tmap)
# Import OA boundaries
liverpool_SP <- readOGR("/Users/sofiagarrick/Documents/COLLEGE/GIS/GIS 3/lab 6/github/urban_analytics/10_Data_Reduction_Geodemographics/data/Liverpool_OA_2011.geojson", "Liverpool_OA_2011")

# Merge lookup
liverpool_SP <- merge(liverpool_SP, lookup, by.x="oa_code",by.y="OA")
m <- tm_shape(liverpool_SP, projection=27700) +
    tm_polygons(col="SUPER", border.col = "black",   palette="Set3",border.alpha = .1, title="Cluster", showNA=FALSE, popup.vars = "Description") +
  tm_layout(legend.position = c("left", "bottom"), frame = FALSE) 

#Create leaflet plot
tmap_leaflet(m)
```

Based on the results of the cluster analysis it seems as though communities in Cluster A tend to be older retired communities living in multi-story flats. Communities in cluster B seem to be middle aged to older individuals living in owned homes with cars with no children or non-dependent children. Communities in cluster C seem to be younger city dwelling families. Communities in cluster D tend to be younger city dwelling individuals who work mostly in manufacturing, utilities, and warehouse services. Communities in cluster E, F, G were mostly characterized by the ethnicity of people living there and were all multicultural. However, cluster E tended to be younger individuals, cluster F older individuals, and clustr G young families. 